Depression and suicidality as costly signals of need

### 2021-04-01 Paper.rmd
Small text edits throughout the paper

# Introduction
Since Durkheim’s pioneering effort to use quantitative social science to explain suicide (cite), anthropologists, psychologists, and sociologists have made varied attempts at explaining depression and suicide. Despite this effort and substantial investment by funding agencies, rates of depression and suicide attempts have remained largely unchanged over the last 100 years (confirm and cite), and there is still substantial uncertainty about why both are as common as they are (cite). Likely due to their high costs and the suffering they cause, the vast majority of these approaches have been built on the assumption that they are the result of internal dysfunction (cite examples). However, a subset of anthropologists and evolutionary theorists have challenged this idea, suggesting that despite the subjective suffering they cause, they may function as a means to limit negative fitness outcomes in times of substantial adversity (Firth + others).

@@

## Connection to Adversity
Likely the most reliable finding when it comes to depression and suicidal behavior is that most cases are closely tied to major negative life events (Hammen, 2005 + others; citations for suicide). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events in both community and clinical samples (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In instances where depression is preceded by negative events, the onset of depression tends to be rapid, often occurring within one month of the event (Kendler et al., 1999 [double check and find others]).

This strong connection between negative life events and depression along with the need and distress that likely accompanies these events has led to the suggestion that this relationship is causal (Kendler et al., 1999). Support for this comes from the fact that the relationship between negative events and depression holds true when looking solely at events outside of one's control, which shows that the connection is unlikely to be driven solely by depressed individuals creating situations where negative events are likely to be common (Hammen, 2005; Kendler et al., 1999). In addition to this, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity and that part of the heritability of depression stems from the heritability of negative events, like divorce and family conflict (Kendler & Baker, 2007; Kendler et al., 1999).

Of the many types of negative life events associated with depression, those which involve social adversity tend to be more strongly associated with depression (Hammen, 2005; Kendler et al., 1999; Mazure, 1998). Like the relationship between negative events as a whole and depression, this relationship is also seen cross-culturally, with the loss of a loved one, threat of divorce, or abusive relationships all increasing one's chances of becoming depressed (Kendler et al., 1999, 1995). Of these events, physical assault has been found to be the strongest predictor of depression in the few studies in which it has been examined (Kendler et al., 1999, 1995) and may have been much more common throughout our evolutionary history than is often thought based on rates of spousal abuse in small-scale societies with little or no policing (Levinson, 1989; Stieglitz et al., 2018).

@@

Although there may be other ways for honest signals evolve (cite), one of the most widely accepted (confirm) means is through costly signaling. The evolution of costly signals is possible whenever the signal in question is comparatively cheaper to produce for those who have the underlying quality compared to those who lack it (Grafen 1990). In these instances, individuals with the quality maximize their fitness at greater levels of investment in the signal compared to those who lack the quality or possess it to a lesser degree (Gangestad and Scheyd 2005; Zahavi 1977). The end result of this is expected to be a signal that either tends to not be worth faking (Gangestad and Scheyd 2005) or is impossible to fake to a large degree due to selection against fakers constraining the physiology behind the signal (Zahavi 1977).

Both depression and suicidal behavior fit this criterion as they are both relatively costly for those not facing adversity than those who are. In the case of depression, the largescale reduction in one's normal routine of fitness enhancing behavior hurts those in need less when one has limited ability to obtain fitness-relevant benefits, with depression likely adding no additional costs in instances of extreme deprivation (Hagen 2003). However, when it comes to suicide, the bulk of the costs associated with suicide attempts are only felt in times that result in death or severe bodily harm. For this reason, the way others view and respond to suicide attempts is likely to be strongly influenced by the probability that the attempt would result in death or severe injury (Rosenthal, 1993). That given, the previously mentioned logic still applies with increased risk of death being relatively costly for those able to enhance their fitness prospects in their day-to-day lives.

@@

In addition to signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support. This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from a suicide. In these instances, depression and suicide attempts are suggested to function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, he or she would need to increase support towards the signaler.

@@

Although negative responses to suicide survivors are more often studied and reported, positive responses have also been found. Increased social support and changes to important relationships are often reported to follow suicide attempts with some indication that these effects may hold long term (Stengel, 1956). For example, a study of 100 women who survived suicide attempts found that in 75 cases individuals gained identifiable benefits through the attempt, with 41 individuals benefiting from reconciliations with others (Lukianowicz, 1971). As for the type of support received, emotional support seems to be more commonly provided than instrumental support based on initial evidence (Frey et al., 2017); however, it is currently unclear how this might change with social context.

@@

## notes: 
add funding info somewhere
update MTurk (Ed email links + mine)
Find a place for the predictions and preregistration 
go back and emphasize conflict and PI in the methods (at least in the pilot studies portion)
add abstract
add conclusion

@@

1,950 participants started the study by clicking the link to Qualtrics (1,213 American and 737 Indian). After removing participants who did not finish the survey, failed attention checks, or moved through the study at an unrealistc pace, our sample dropped to N = 1,240 (937 American and 303 Indian). Basic summary statistics on the remaining N = 1,240 participants (Male = 759, Female = 479) can be found in table XX.

@@

Power analyses prior to data collection showed that high proportions of participants scoring extremely high on belief or likelihood of helping at T1 would result in low power for this study's design, a problem which we appeared to largely avoid (see figure XX).

@@

As a whole, our predictions for the effect of signal cost on belief and likelihood of helping T2 were met, with costlier signals eliciting greater pro-signaler responses. There was also support for an interaction between signal and T1Belief or T1Action, although the effect was not large (See figure 2). However, the effect of signal cost varied by vignette, with signal type having the greatest effect in the "basketball coach" and "brother-in-law" vignettes. When ordered according to cost, the greatest jump in support tended to occur between crying and mild-depression, with the next largest jump tending to occur between depression and suicide attempt. Overall, there was little difference between the effect of mild depression and the effect of depression.

@@

We also examined the effect of signaling on how participants perceived the signaler's emotional state. To do this, we created four new variables: LowMood (T1 and T2) and Manipulative (T1 and T2). This was done by adding the dummy variables for whether a participant thought the signaler was depressed, distressed or sad for the LowMood variables and by adding the dummy variables for whether they thought the particpant was Devious or Jealous for the Manipulative variables. In all instances, 0 indicated a participant did not select the emotion, while 1 indicated selection.

We then plotted the mean change seen in the data in addition to the change seen in 500 bootstrap resamples (see figure 6).

@@

Although this study was not designed to examine whether the costs of signaling would be made up for by the increased support stemming from depression and suicide attempts directly, the results are encouraging (reword). In general, the costs required to honestly convey one's qualities should be expected to increase with the potential payoffs of deception (cite), with the vignettes in this study emphasizing the existence of private information and sizable conflicts of interest to increase the perceived risk of being taken advantage of. In similar real-world scenarios, this difficulty of signaling honest need would coincide with the substantial fitness-relevant benefits of preventing future assaults or being able to marry the partner of one's choice. These benefits coupled with the fact (too strong?) that depression and suicide attempts become relatively less costly with increased need (Hagen 2003), suggest that the increased support seen in this study may be worth the costs required to signal need more often than is often appreciated when counting the entire costs of depression as the signal rather than only the upregulation of symptoms for a communicative purpose.

>>>>

# Introduction
Since Durkheim’s pioneering effort to use quantitative social science to explain suicide (cite), anthropologists, psychologists, and sociologists have made varied attempts at better understanding depression and suicide. Despite this effort and substantial investment by funding agencies, rates of depression and suicide attempts have remained largely unchanged over the last 100 years (confirm and cite), and there is still substantial uncertainty about why both are as common as they are (cite). Likely due to  the suffering they cause, the vast majority of these approaches have been built on the assumption that they are the result of internal dysfunction (cite examples). However, a subset of anthropologists and evolutionary theorists have challenged this idea, suggesting that despite the subjective suffering they cause, they may function as a means to limit negative fitness outcomes in times of substantial adversity (Firth + others).

@@

## Connection to Adversity
Likely the most reliable finding when it comes to depression and suicidal behavior is that most cases are closely tied to major negative life events (Hammen, 2005 + others; citations for suicide). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events in both community and clinical samples (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In instances where depression is preceded by negative events, it's onset tends to quickly follow, often occurring within one month of the event (Kendler et al., 1999 [double check and find others]).

This strong connection between negative life events and depression along with the adversity and distress that likely accompanies these events has led to the suggestion that this relationship is causal (Kendler et al., 1999). Support for this comes from the fact that the relationship between negative events and depression holds true when looking solely at events outside of one's control, showing that the connection is unlikely to be driven solely by depressed individuals creating situations where negative events are likely to be common (Hammen, 2005; Kendler et al., 1999). In addition to this, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity and that part of the heritability of depression stems from the heritability of negative events, like divorce and family conflict (Kendler & Baker, 2007; Kendler et al., 1999).

Of the many types of negative life events associated with depression, those which involve social adversity tend to be more strongly associated with depression (Hammen, 2005; Kendler et al., 1999; Mazure, 1998). Like the relationship between negative events as a whole and depression, this relationship is also seen cross-culturally, with the loss of a loved one, threat of divorce, or abusive relationships all increasing one's chances of becoming depressed (Kendler et al., 1999, 1995 [check if this fits the cross-cultural claim]). Of these events, physical assault has been found to be the strongest predictor of depression in the few studies in which it has been examined (Kendler et al., 1999, 1995) and may have been much more common throughout our evolutionary history than is often appreciated based on rates of spousal abuse in small-scale societies with little or no policing (Levinson, 1989; Stieglitz et al., 2018).

@@

Although there may be other ways for honest signals evolve (cite), one of the most widely accepted (is this fair?) means is through costly signaling. The evolution of costly signals is possible whenever the signal in question is comparatively cheaper to produce for those who have the underlying quality compared to those who lack it (Grafen 1990). In these instances, individuals with the quality maximize their fitness at greater levels of investment in the signal compared to those who lack the quality or possess it to a lesser degree (Gangestad and Scheyd 2005; Zahavi 1977). The end result of this is expected to be a signal that either tends to not be worth faking (Gangestad and Scheyd 2005) or is impossible to fake to a large degree due to selection against fakers constraining the physiology behind the signal (Zahavi 1977).

Both depression and suicidal behavior fit this criterion as they are both relatively costly for those not facing adversity than those who are. In the case of depression, the largescale reduction in one's normal routine of fitness enhancing behavior hurts those in need less when one has limited ability to obtain fitness-relevant benefits, with depression likely adding no additional costs in instances of extreme deprivation (Hagen 2003). However, when it comes to suicide, the bulk of the costs associated with suicide attempts are only felt in times that result in death or severe bodily harm. For this reason, the way others view and respond to suicide attempts is likely to be strongly influenced by the probability that the attempt would result in death or severe injury (Rosenthal, 1993). That given, the previously mentioned logic still applies with increased risk of death being relatively costly for those able to advance their fitness prospects in their day-to-day lives.

@@

In addition to signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support (Hagen 2003). This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from a suicide. In these instances, depression and suicide attempts are suggested to function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, he or she would need to increase support towards the signaler.

@@

Although negative responses to suicide survivors are more often studied and reported, positive responses have also been found. Increased social support and changes to important relationships have been reported to follow suicide attempts with some indication that these effects may hold long term (Stengel, 1956). For example, a study of 100 women who survived suicide attempts found that in 75 cases individuals gained identifiable benefits through the attempt, with 41 individuals benefiting from reconciliations with others (Lukianowicz, 1971). As for the type of support received, emotional support seems to be more commonly provided than instrumental support based on initial evidence (Frey et al., 2017); however, it is currently unclear how this might change with social context.

@@

## notes: 
add funding info somewhere
Find a place for the predictions and preregistration 
go back and emphasize conflict and PI in the methods (at least in the pilot studies portion)
add abstract
add conclusion

@@

1,950 participants started the study by clicking the link to Qualtrics (1,213 American and 737 Indian). After removing participants who did not finish the survey, failed attention checks, or moved through the study at an unrealistic pace, our sample dropped to N = 1,240 (937 American and 303 Indian). Basic summary statistics on the remaining N = 1,240 participants (Male = 759, Female = 479) can be found in table XX.

@@

Power analyses prior to data collection showed that high proportions of participants scoring extremely high on belief or likelihood of helping at T1 would result in low power for this study's design, a problem which we appeared to largely avoid (word better) (see figure XX).

@@

As a whole, our predictions for the effect of signal cost on belief and likelihood of helping T2 were met, with costlier signals eliciting greater pro-signaler responses (reword). There was also support for an interaction between signal and T1Belief or T1Action, although the effect was not large (See figure 2). However, the effect of signal cost varied by vignette, with signal type having the greatest effect in the "basketball coach" and "brother-in-law" vignettes. When ordered according to cost, the greatest jump in support tended to occur between crying and mild-depression, with the next largest jump tending to occur between depression and suicide attempt. Overall, there was little difference between the effect of mild depression and the effect of depression.

@@

We also examined the effect of signaling on how participants perceived the signaler's emotional state. To do this, we created four new variables: LowMood (T1 and T2) and Manipulative (T1 and T2). This was done by adding the dummy variables for whether a participant thought the signaler was depressed, distressed or sad for the LowMood variables and by adding the dummy variables for whether they thought the participant was Devious or Jealous for the Manipulative variables. For each original emotion variable, 0 indicated a participant did not select the emotion, while 1 indicated selection.

We then plotted the mean change in the LowMood and Manipulative variables seen in the data in addition to the change seen in 500 bootstrap resamples (see figure 6).

@@

Although this study was not designed to examine whether the costs of signaling would be made up for by the increased support stemming from depression and suicide attempts directly, the results are suggestive (reword?). In general, the costs required to honestly convey one's qualities should be expected to increase with the potential payoffs of deception (cite), with the vignettes in this study emphasizing the existence of private information and sizable conflicts of interest to increase the perceived risk of being taken advantage of. In similar real-world scenarios, this difficulty of signaling honest need would coincide with the substantial fitness-relevant benefits of preventing future assaults or being able to marry the partner of one's choice. These benefits coupled with the fact (too strong?) that depression and suicide attempts become relatively less costly with increased need (Hagen 2003), suggest that the increased support seen in this study may be worth the costs required to signal need more often than is often appreciated when counting the entire costs of depression as the signal rather than only the upregulation of symptoms for a communicative purpose.
